---
title: "MiniProject1"
author: "Grace Rubio, Oyinda Sanni, Mia Mason"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

**Important Note** All code needs to have an answer to the "Where did u
get ur code" question

# Preparing the Data

## Load in packages

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(tidymodels)
```

## Importing Data

1.  Import the data

```{r}
# Response Variable
life_exp <-read_csv("Data/lex.csv") 

# Predictor Variables
CO2_emissions <-read_csv("Data/co2_cons.csv")
gdp <-read_csv("Data/gdp_pcap.csv")
humanDevIndx <-read_csv("Data/hdi_human_development_index.csv")
poverty<-read_csv("Data/number_of_people_in_poverty.csv")
gapminder_2025<-read_csv("Data/gapminder_2025.csv") # QUALITATIVE (regions) 
```

## Wrangling Data

TODO: 1. Combine all the data into a large dataset

*NOTE:* We will use the last recorded year for the datasets that do not
have a 2025 entry.

```{r}
data_lst<- list(CO2_emissions, gdp, humanDevIndx, poverty, gapminder_2025)

Full_2025<-life_exp

for (dataset in data_lst){
 Full_2025 <- Full_2025|>
  full_join(dataset, by="geo") 
}

Full_2025<-Full_2025|>
  select(geo,name.x,`2025.x`, # keep life_exp 2025 columns
         `2022.y`, # keep 2025 co2 column
         `2025.y`,  # keep 2025 gdp column
         `2023.x.x`,  # keep 2025 humanDevIdnx column
         `2025`,     # keep 2025 poverty column
         world_4region # keep 2025 regions column
         )|>
  rename(name = name.x, #rename columns for clarity
         life_Exp=`2025.x`,
         CO2=`2022.y`,
         GDP=`2025.y`,
         Poverty = `2025`,
         HumDevIndx= `2023.x.x`
         )

```

# Variable Report

TODO: 1. Identify each variable 2. Type of each variable 3. Information
about the types of values the variable can take 4. Amount of missing
data within the variable 5. Check if cleaning is needed and clean the
data.

## Data Cleaning

For the most part the `Full_2025` table is tidy. The main issue with the
table is with the missing values in a few of the variables. As such, we
will clean the table by removing all missing entries.

*Note:* Below, we further identify issues with each individual variable
specifically. This way, we can explicitly distinguish if/why the
cleaning performed in this section was necessary.

```{r}
# Clean the Data 
Full_2025<-Full_2025|>
  drop_na()
```

## CO2 Emissions

*Type:* This variable type is quantitative.

*Info:* This variable is the carbondioxide total (consumption based).
This variable can take numeric or NAN values. The CO2 emissions column
is measured in million tonnes of CO2. There are currently two missing
entries in the variable, so this column does require some cleaning.

### Summary Statistics

TODO: Quantitative variables: include basic summary stats like mean,
median, standard deviation, min, max, etc. 2a. Create an appropriate
plot to show distribution of values 3. Look up country involved for any
extreme outliers

```{r}
#| echo: true
#| fig-align: center

# SUMMARY
CO2_summary <- Full_2025|>
  summarize(avg_CO2 = mean(CO2), 
                        sd_CO2 = sd(CO2), 
                        min_CO2 = min(CO2), 
                        max_CO2 = max(CO2), 
                        mdn_CO2 = median(CO2), 
                        iqr = IQR(CO2),
                        sample_size = n())
CO2_summary

# PLOT
Full_2025|>
  ggplot(mapping=aes(x = CO2)) +
  geom_histogram(binwidth = 300, color = '#ACC7B4', fill = "#331B3F") +
  labs(x = "Total CO2 Emissions (Million Tonnes)",
       y = "Count",
       title = "Count of Country's Total CO2 Emission")+
  # Add markers to identify outliers
  geom_point(data = Full_2025[Full_2025$CO2 > 2000,],aes(x = CO2, y = 0), 
             color = "red", shape = 1, size = 12)

```

### Variable Descriptions

We can see that the distribution of total CO2 emissions across the
listed countries is right-skewed with the majority of countries falling
before the 1 billion tonnes mark. There are two main outliers of this
variable (India and the US) that likely have an impact on our summary
results, so that should be taken into account with later analysis. Given
the results of our summary for this variable, we can see that the
average CO2 emission per country is about 137.15 million tonnes. By
comparing the mean to our median (17.7 million tonnes), the
right-skewness of the data becomes more apparent as the difference
between the average CO2 emission and the middle value further suggests
the outliers are pulling the mean up. The IQR is helpful for this skewed
distribution, as it tells us that the middle 50% of countries have CO₂
emissions within a range of about 70 million tonnes.

## GDP per Capita

*Type:* This variable type is quantitative.

*Info:*

### Summary Statistics

TODO: Quantitative variables: include basic summary stats like mean,
median, standard deviation, min, max, etc. 2a. Create an appropriate
plot to show distribution of values 3. Look up country involved for any
extreme outliers

```{r}

```

### Variable Descriptions

TODO: 1. Write a few sentences for each variable to describe what we
have learned about the variable. Incorporate information from the
Summary Statistics

## Human Development Index

*Type:* This variable type is quantitative.

*Info:*

### Summary Statistics

TODO: Quantitative variables: include basic summary stats like mean,
median, standard deviation, min, max, etc. 2a. Create an appropriate
plot to show distribution of values 3. Look up country involved for any
extreme outliers

```{r}

```

### Variable Descriptions

TODO: 1. Write a few sentences for each variable to describe what we
have learned about the variable. Incorporate information from the
Summary Statistics

## People in Poverty

*Type:* This variable type is quantitative.

*Info:*

### Summary Statistics

TODO: Quantitative variables: include basic summary stats like mean,
median, standard deviation, min, max, etc. 2a. Create an appropriate
plot to show distribution of values 3. Look up country involved for any
extreme outliers

```{r}

```

### Variable Descriptions

TODO: 1. Write a few sentences for each variable to describe what we
have learned about the variable. Incorporate information from the
Summary Statistics

## Region

*Type:* This variable type is categorical.

*Info:*

### Summary Statistics

TODO: 1. Categorical variables: create a table of possible values and
include counts or proportions. 1a. Create an appropriate plot to show
distribution of values

```{r}

```

### Variable Descriptions

TODO: 1. Write a few sentences for each variable to describe what we
have learned about the variable. Incorporate information from the
Summary Statistics

# Linear Models

TODO: 1. Make at least 3 linear models (1 simple, 2 multivariate) 2.
Evaluate the models for LINE(Q-Q plots needed), p-values ,R2, and
adjusted R2, RSE, MSE, RMSE

```{r}

```

# Comparing Models

TODO: 1. Which model is the best?

# Questions About the Data

TODO: 1. Ask a question (Should explore relationship between at least 2
variables in the dataset) 2. Answer the question 3. Create a graph/plot
that helps answer the question

```{r}

```
