---
title: "MiniProject1"
author: "Grace Rubio, Oyinda Sanni, Mia Mason"
format: pdf 
editor:    
  markdown:      
    wrap: 72 
---

**Important Note** All code needs to have an answer to the "Where did u
get ur code" question

# Preparing the Data

## Load in packages

```{r}
#| label: setup 
#| include: false
#| library(tidyverse)
#| library(tidymodels)
```

## Importing Data

1.  Import the data

```{r}
# Response Variable
life_exp <-read_csv("Data/lex.csv")   
# Predictor Variables 
CO2_emissions <-read_csv("Data/co2_cons.csv")
gdp <-read_csv("Data/gdp_pcap.csv")
humanDevIndx <-read_csv("Data/hdi_human_development_index.csv")
poverty<-read_csv("Data/number_of_people_in_poverty.csv")
gapminder_2025<-read_csv("Data/gapminder_2025.csv") # QUALITATIVE (regions) 
```

## Wrangling Data

TODO: 1. Combine all the data into a large dataset

*NOTE:* We will use the last recorded year for the datasets that do not
have a 2025 entry.

```{r}
data_lst<- list(CO2_emissions, gdp, humanDevIndx, poverty, gapminder_2025)

Full_2025<-life_exp 
for (dataset in data_lst){  
  Full_2025 <- Full_2025|>
    full_join(dataset, by="geo")
}

Full_2025<-Full_2025|>
  select(geo,name.x,`2025.x`, # keep life_exp 2025 columns
         `2022.y`, # keep 2025 co2 column
         `2025.y`,  # keep 2025 gdp column
         `2023.x.x`,  # keep 2025 humanDevIdnx column
         `2025`,     # keep 2025 poverty column
         world_4region # keep 2025 regions column
         )|>
  rename(name = name.x, #rename columns for clarity
         life_Exp=`2025.x`,
         CO2=`2022.y`,
         GDP=`2025.y`,
         Poverty = `2025`,
         HumDevIndx= `2023.x.x`
         ) 
```

# Variable Report

TODO: 1. Identify each variable 2. Type of each variable 3. Information
about the types of values the variable can take 4. Amount of missing
data within the variable 5. Check if cleaning is needed and clean the
data.

## Data Cleaning

For the most part the `Full_2025` table is tidy. The main issue with the
table is with the missing values in a few of the variables. As such, we
will clean the table by removing all missing entries.

*Note:* Below, we further identify issues with each individual variable
specifically. This way, we can explicitly distinguish if/why the
cleaning performed in this section was necessary.

```{r}
# Clean the Data
Full_2025<-Full_2025|>   
  drop_na()
```

## CO2 Emissions

*Type:* This variable type is quantitative.

*Info:* This variable is the carbondioxide total (consumption based).
This variable can take numeric or NAN values. The CO2 emissions column
is measured in million tonnes of CO2. There are currently two missing
entries in the variable, so this column does require some cleaning.

### Summary Statistics

```{r}
#| echo: true
#| fig-align: center  

# SUMMARY
CO2_summary <- Full_2025|>
  summarize(avg_CO2 = mean(CO2),
            sd_CO2 = sd(CO2),
            min_CO2 = min(CO2),
            max_CO2 = max(CO2),
            mdn_CO2 = median(CO2),
            iqr = IQR(CO2),
            sample_size = n())
CO2_summary

# PLOT
Full_2025|>
  ggplot(mapping=aes(x = CO2)) +
  geom_histogram(binwidth = 300, color = '#ACC7B4', fill = "#331B3F") +
  labs(x = "Total CO2 Emissions (Million Tonnes)",
       y = "Count",
       title = "Count of Country's Total CO2 Emission")+

  # Add markers to identify outliers
  geom_point(data = Full_2025[Full_2025$CO2 > 2000,],
             aes(x = CO2, y = 0),
             color = "red", shape = 1, size = 12) 
```

### Variable Descriptions

We can see that the distribution of total CO2 emissions across the
listed countries is right-skewed with the majority of countries falling
before the 1 billion tonnes mark. There are two main outliers of this
variable (India and the US) that likely have an impact on our summary
results, so that should be taken into account with later analysis. Given
the results of our summary for this variable, we can see that the
average CO2 emission per country is about 137.15 million tonnes. By
comparing the mean to our median (17.7 million tonnes), the
right-skewness of the data becomes more apparent as the difference
between the average CO2 emission and the middle value further suggests
the outliers are pulling the mean up. The IQR is helpful for this skewed
distribution, as it tells us that the middle 50% of countries have CO₂
emissions within a range of about 70 million tonnes.

## GDP per Capita

*Type:* This variable type is quantitative.

*Info:* GDP per capita is the gross domestic product which is the total
moetary value of all goods and services produced in a country in a given
year. The values in this dataset are stored as float values. GDP per
capita is recorded in US dollars, specifically PPP\$ or purchasing power
parity, and values are adjusted for inflation.

### Summary Statistics

```{r}
#| echo: true
#| fig-align: center

# SUMMARY
GDP_summary <- Full_2025|>
  summarize(avg_GDP = mean(GDP),
            sd_GDP = sd(GDP),
            min_GDP = min(GDP),
            max_GDP = max(GDP),
            mdn_GDP = median(GDP),
            iqr = IQR(GDP),
            sample_size = n())
GDP_summary

# PLOT
Full_2025|>
  ggplot(mapping=aes(x = GDP)) +
  geom_histogram(binwidth = 3000, color = '#ACC7B4', fill = "#331B3F") +
  labs(x = "Total GDP per Capita (Dollars)",
       y = "Count",
       title = "Count of Country's Total GDP per Capita")+
# Add markers to identify outliers
  geom_point(data = Full_2025[Full_2025$GDP > 90000,],
             aes(x = GDP, y = 0),
             color = "red", shape = 1, size = 12)
```

### Variable Descriptions

The distribution of GDP per capita across the listed countries in the
Full_2025 dataset is concentrated on the right side of the graph, with a
majority of it being below \$25,000. There are four outliers which reach
up to around \$140,000, causing the mean to be higher in the summary.
Given the summary, the average GDP per capita is \~\$24,859, which
comparing to the median of \$15,343 highlights the right-sided
concentration. The IQR of \$32,353 provides a better sense of spread for
this distribution, indicating that the middle 50% of countries fall
within a range of roughly \$32,353.

## Human Development Index

*Type:* This variable type is quantitative.

*Info:* The HumDevIndx variable is the measure of overall human
well-being and development of a country. It combines three key
dimensions of human development like health or longevity, education, and
standard of living. These area measured by life expectancy at birth,
mean years of schooling for adults and expected years of schooling for
children, and the gross national income per capita respectively. The
variable ranges from 0 to 1, where 1 indicates higher levels of human
development, and is a float type.

### Summary Statistics

```{r}
#| echo: true
#| fig-align: center

# SUMMARY
HumDevIndx_summary <- Full_2025|>
  summarize(avg_HDI = mean(HumDevIndx),
            sd_HDI = sd(HumDevIndx),
            min_HDI = min(HumDevIndx),
            max_HDI = max(HumDevIndx),
            mdn_HDI = median(HumDevIndx),
            iqr = IQR(HumDevIndx),
            sample_size = n())
HumDevIndx_summary

# PLOT
Full_2025|>
  ggplot(mapping=aes(x = HumDevIndx)) +
  geom_histogram(binwidth = 0.02, color = '#ACC7B4', fill = "#331B3F") +
  labs(x = "Total Human Development Index",
       y = "Count",
       title = "Count of Country's Total Human Development Index")
```

### Variable Descriptions

The distribution of the HDI scores across the countries is concentrated
to the left, with the majority of them having an HDI score above 0.7.
There are no extreme outliers in the distribution due to the variable
being bound to a scale with a minimum of 0.388 and a maximum of 0.972.
Given the results of the summary for the variable, the average HDI score
across countries is approximately 0.734. Comparing the mean to the
median of 0.75, the slight left skew becomes apparent, as the middle
value sits slightly higher than the average due to the tail of lower HDI
scores bringing the mean down. The IQR shows that the middle 50% of
countries fall within a narrow range of about 0.236.

## People in Poverty

*Type:* This variable type is quantitative.

*Info:* The poverty variable records the estimated population in a
country whose daily income or consumption falls before the international
extreme poverty threshold. The unit of measure is number of people.

### Summary Statistics

```{r}
#| echo: true
#| fig-align: center

# SUMMARY
Poverty_summary <- Full_2025|>
  summarize(avg_Poverty = mean(Poverty),
            sd_Poverty = sd(Poverty),
            min_Poverty = min(Poverty),
            max_Poverty = max(Poverty),
            mdn_Poverty = median(Poverty),
            iqr = IQR(Poverty),
            sample_size = n()) 
Poverty_summary

# PLOT
Full_2025|>   
  ggplot(mapping=aes(x = Poverty)) +
  geom_histogram(binwidth = 3000000, color = '#ACC7B4', fill = "#331B3F") +
  labs(x = "Total Poverty Population (Millions of People)",
       y = "Count",
       title = "Count of Country's Total Poverty Population")+

# Add markers to identify outliers
  geom_point(data = Full_2025[Full_2025$Poverty > 40000000,],
             aes(x = Poverty, y = 0),
             color = "red", shape = 1, size = 12)
```

### Variable Descriptions

The distribution of the total poverty population across the countries is
very right-skewed, with the majority of them being clustered below 10
million. There are four distinct outliers to the right with the maximum
reaching a population as high as 95,605,313 people. This maximum value
strongly influences the results of the summary table and should be taken
into account for the analysis. From the summary results, the average
poverty population per country is 4,615,385 people, and a very high
standard deviation of 12,475,121. Comparing the mean to the median of
182,163 people, the extreme right skew is very evident and shows how the
few high-poverty countries inflate the mean. The IQR shoes that the
middle 50% of countries have a poverty population spanning a range of
about 2.42 million people.

## Region

*Type:* This variable type is categorical.

*Info:* The region variable stores information about the continent in
which the country is located in. The variable can be one of the four
following values: americas (includes north and south), africa, asia, and
europe.

### Summary Statistics

```{r}
#| echo: true
#| fig-align: center

# SUMMARY
region_summary <- Full_2025|>
  group_by(world_4region) |>
  summarize(count = n())
region_summary  

# PLOT 
Full_2025|>
  ggplot(mapping=aes(x = world_4region)) +
  geom_bar(color = '#ACC7B4', fill = "#331B3F") +
  labs(x = "World Regions",
       y = "Count", 
       title = "Distribution of Countries Across World Regions")
```

### Variable Descriptions

Africa and Asia are the most represented regions out of the countries in
the dataset with their counts being 53 and 52 respectively. The Americas
are the least represented world region being the minimum at 34 total
countries.

# Linear Models

TODO: 1. Make at least 3 linear models (1 simple, 2 multivariate) 2.
Evaluate the models for LINE(Q-Q plots needed), p-values ,R2, and
adjusted R2, RSE, MSE, RMSE

## GDP per Capita: Simple Linear Model

```{r}
lifegdp=Full_2025|>
  lm(life_Exp~GDP, data= _) 
lifegdp
```

### Interpreting the Slope

*Equation:* $\hat{y}_{LifeExp}=0.6851+0.0001935x_{GDP}$

*Slope:* 0.0001935

*Interpretation:* The slope of this model tells us that For every \$1
increase in gdp per capita there will be a 0.0001935 year increase in
predicted life expectancy. In other words, the predicted life expectancy
of a country increases by 1.9 years for every \$10,000 increase in GDP.

### Evaluating the Model

```{r}
plot(lifegdp, which = 1)
plot(lifegdp, which = 2) 
```

L - Bad. The residuals vs fitted plot shows us a poor linearity between
the variables. We can see this because the points do not cluster/ evenly
distribute around zero. We can aslo see the line curve downwards as the
fitted values increase.

I - Not good. There are patterns in the geographic location/region of
the data. As such, we cannot assume independence.

N - Ok. The Q-Q plot shows the data lying closely along the line for
(the most part). While there is some minor divergence (especially
towards the end), we can still say that it is reasonable to assume
general normal distribution among the variables.

E - Bad. The variance declines as fitted value increases (i.e., the
variance is NOT constant).

### Model Summary

```{r}
summary(lifegdp)
```

#### p-value

**p-value=2.2e-16**

Since the p-value is very small (threshold: $p<=0.05$), we can say that
the slope is significantly different from 0.

#### $R^2$

$R^2 = 0.5041$

50.41% of the variation of life expectancy is explained by the linear
model including gdp per capita.

#### RSE

$RSE = 4.93$

The "average" distance of the life expectancy from its predicted value
is about 5 years.

## Poverty & CO2 Emissions: Multivariate Model

```{r}
firstmultiv=Full_2025|>
  lm(life_Exp~CO2+Poverty, data= _) 
firstmultiv
```

### Interpreting the Slope

*Equation:* $x$

*Slope:*

*Interpretation:*

### Evaluating the Model

```{r}
plot(firstmultiv, which = 1)
plot(firstmultiv, which = 2) 
```

L -

I -

N -

E -

### Model Summary

```{r}
summary(firstmultiv)
```

#### p-value(s)

#### $R^2$

#### Adjusted $R^2$

#### RSE

## Human Development Index & GDP & Region: Multivariate Model

```{r}
secondmultiv=Full_2025|>
  lm(life_Exp~HumDevIndx+GDP+world_4region, data= _) 
secondmultiv
```

### Interpreting the Slope

*Equation:* $x$

*Slope:*

*Interpretation:*

### Evaluating the Model

```{r}
plot(secondmultiv, which = 1)
plot(secondmultiv, which = 2) 
```

L -

I -

N -

E -

### Model Summary

```{r}
summary(secondmultiv)
```

#### p-value(s)

#### $R^2$

#### Adjusted $R^2$

#### RSE

# Comparing Models

TODO: 1. Which model is the best?

# Questions About the Data

TODO: 1. Ask a question (Should explore relationship between at least 2
variables in the dataset) 2. Answer the question 3. Create a graph/plot
that helps answer the question

*Question:*

```{r}
```

*Answer:*
