---
title: "MiniProject1"
author: "Grace Rubio, Oyinda Sanni, Mia Mason"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

**Important Note** All code needs to have an answer to the "Where did u
get ur code" question

# Preparing the Data

## Load in packages

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(tidymodels)
```

## Importing Data

1.  Import the data

```{r}
# Response Variable
life_exp <-read_csv("Data/lex.csv") 

# Predictor Variables
CO2_emissions <-read_csv("Data/co2_cons.csv")
gdp <-read_csv("Data/gdp_pcap.csv")
humanDevIndx <-read_csv("Data/hdi_human_development_index.csv")
poverty<-read_csv("Data/number_of_people_in_poverty.csv")
gapminder_2025<-read_csv("Data/gapminder_2025.csv") # QUALITATIVE (regions) 
```

## Wrangling Data

TODO: 1. Combine all the data into a large dataset

*NOTE:* We will use the last recorded year for the datasets that do not
have a 2025 entry.

```{r}
data_lst<- list(CO2_emissions, gdp, humanDevIndx, poverty, gapminder_2025)

Full_2025<-life_exp

for (dataset in data_lst){
 Full_2025 <- Full_2025|>
  full_join(dataset, by="geo") 
}

Full_2025<-Full_2025|>
  select(geo,name.x,`2025.x`, # keep life_exp 2025 columns
         `2022.y`, # keep 2025 co2 column
         `2025.y`,  # keep 2025 gdp column
         `2023.x.x`,  # keep 2025 humanDevIdnx column
         `2025`,     # keep 2025 poverty column
         world_4region # keep 2025 regions column
         )|>
  rename(name = name.x, #rename columns for clarity
         life_Exp=`2025.x`,
         CO2=`2022.y`,
         GDP=`2025.y`,
         Poverty = `2025`,
         HumDevIndx= `2023.x.x`
         )

```

# Variable Report

TODO: 1. Identify each variable 2. Type of each variable 3. Information
about the types of values the variable can take 4. Amount of missing
data within the variable 5. Check if cleaning is needed and clean the
data.

## Data Cleaning

For the most part the `Full_2025` table is tidy. The main issue with the
table is with the missing values in a few of the variables. As such, we
will clean the table by removing all missing entries.

*Note:* Below, we further identify issues with each individual variable
specifically. This way, we can explicitly distinguish if/why the
cleaning performed in this section was necessary.

```{r}
# Clean the Data 
Full_2025<-Full_2025|>
  drop_na()
```

## CO2 Emissions

*Type:* This variable type is quantitative.

*Info:* This variable is the carbondioxide total (consumption based).
This variable can take numeric or NAN values. The CO2 emissions column
is measured in million tonnes of CO2. There are currently two missing
entries in the variable, so this column does require some cleaning.

### Summary Statistics

TODO: Quantitative variables: include basic summary stats like mean,
median, standard deviation, min, max, etc. 2a. Create an appropriate
plot to show distribution of values 3. Look up country involved for any
extreme outliers

```{r}
#| echo: true
#| fig-align: center

# SUMMARY
CO2_summary <- Full_2025|>
  summarize(avg_CO2 = mean(CO2), 
                        sd_CO2 = sd(CO2), 
                        min_CO2 = min(CO2), 
                        max_CO2 = max(CO2), 
                        mdn_CO2 = median(CO2), 
                        iqr = IQR(CO2),
                        sample_size = n())
CO2_summary

# PLOT
Full_2025|>
  ggplot(mapping=aes(x = CO2)) +
  geom_histogram(binwidth = 300, color = '#ACC7B4', fill = "#331B3F") +
  labs(x = "Total CO2 Emissions (Million Tonnes)",
       y = "Count",
       title = "Count of Country's Total CO2 Emission")+
  # Add markers to identify outliers
  geom_point(data = Full_2025[Full_2025$CO2 > 2000,],aes(x = CO2, y = 0), 
             color = "red", shape = 1, size = 12)

```

### Variable Descriptions

We can see that the distribution of total CO2 emissions across the
listed countries is right-skewed with the majority of countries falling
before the 1 billion tonnes mark. There are two main outliers of this
variable (India and the US) that likely have an impact on our summary
results, so that should be taken into account with later analysis. Given
the results of our summary for this variable, we can see that the
average CO2 emission per country is about 137.15 million tonnes. By
comparing the mean to our median (17.7 million tonnes), the
right-skewness of the data becomes more apparent as the difference
between the average CO2 emission and the middle value further suggests
the outliers are pulling the mean up. The IQR is helpful for this skewed
distribution, as it tells us that the middle 50% of countries have CO₂
emissions within a range of about 70 million tonnes.

## GDP per Capita

*Type:* This variable type is quantitative.

*Info:*

### Summary Statistics

TODO: Quantitative variables: include basic summary stats like mean,
median, standard deviation, min, max, etc. 2a. Create an appropriate
plot to show distribution of values 3. Look up country involved for any
extreme outliers

```{r}

```

### Variable Descriptions

TODO: 1. Write a few sentences for each variable to describe what we
have learned about the variable. Incorporate information from the
Summary Statistics

## Human Development Index

*Type:* This variable type is quantitative.

*Info:*

### Summary Statistics

TODO: Quantitative variables: include basic summary stats like mean,
median, standard deviation, min, max, etc. 2a. Create an appropriate
plot to show distribution of values 3. Look up country involved for any
extreme outliers

```{r}

```

### Variable Descriptions

TODO: 1. Write a few sentences for each variable to describe what we
have learned about the variable. Incorporate information from the
Summary Statistics

## People in Poverty

*Type:* This variable type is quantitative.

*Info:*

### Summary Statistics

TODO: Quantitative variables: include basic summary stats like mean,
median, standard deviation, min, max, etc. 2a. Create an appropriate
plot to show distribution of values 3. Look up country involved for any
extreme outliers

```{r}

```

### Variable Descriptions

TODO: 1. Write a few sentences for each variable to describe what we
have learned about the variable. Incorporate information from the
Summary Statistics

## Region

*Type:* This variable type is categorical.

*Info:*

### Summary Statistics

TODO: 1. Categorical variables: create a table of possible values and
include counts or proportions. 1a. Create an appropriate plot to show
distribution of values

```{r}

```

### Variable Descriptions

TODO: 1. Write a few sentences for each variable to describe what we
have learned about the variable. Incorporate information from the
Summary Statistics

# Linear Models

TODO: 1. Make at least 3 linear models (1 simple, 2 multivariate) 2.
Evaluate the models for LINE(Q-Q plots needed), p-values ,R2, and
adjusted R2, RSE, MSE, RMSE

## GDP per Capita: Simple Linear Model

```{r}
lifegdp=Full_2025|>
  lm(life_Exp~GDP, data= _) 
lifegdp
```

### Interpreting the Slope

*Equation:* $\hat{y}_{LifeExp}=68.51+0.0001935x_{GDP}$

*Slope:* 0.0001935

*Interpretation:* The slope of this model tells us that For every \$1
increase in gdp per capita there will be a 0.0001935 year increase in
predicted life expectancy. In other words, the predicted
`life expectancy` of a country increases by 1.9 years for every \$10,000
increase in `GDP`.

### Evaluating the Model

```{r}
plot(lifegdp, which = 1)
plot(lifegdp, which = 2) 
```

L - Bad. The residuals vs fitted plot shows us a poor linearity between
the variables. We can see this because the points do not cluster/ evenly
distribute around zero. We can aslo see the line curve downwards as the
fitted values increase.

I - Not good. There are patterns in the geographic location/region of
the data. As such, we cannot assume independence.

N - Okay. The Q-Q plot shows the data lying closely along the line for
(the most part). While there is some minor divergence (especially
towards the end), we can still say that it is reasonable to assume
general normal distribution among the variables.

E - Bad. The variance declines as fitted value increases (i.e., the
variance is NOT constant).

### Model Summary

```{r}
summary(lifegdp)
```

#### p-value

$p-value=2.2e-16$

Since the p-value is very small (threshold: $p<=0.05$), we can say that
the slope is significantly different from 0. Therefore it is
*statistically significant*.

#### $R^2$

$R^2 = 0.5041$

50.41% of the variation of life expectancy is explained by the linear
model including gdp per capita.

#### RSE

$RSE = 4.93$

The "average" distance of the life expectancy from its predicted value
is about 5 years.

## Poverty & Human Development Index: Multivariate Model

```{r}
firstmultiv=Full_2025|>
  lm(life_Exp~HumDevIndx+Poverty, data= _) 
firstmultiv
```

### Interpreting the Slope

*Equation:*
$\hat{y}_{LifeExp}=44.07+39.85x_{HumDevIndx}+0.000000003099x_{Poverty}$

*Slope:*

**Human Dev Index:** 39.85

**Poverty:** 0.000000003099

*Interpretation:*

**Human Dev Index:** The predicted `life expectancy` increases by about
4 years with every 0.1 unit increase in the `Human Development Index`,
when the number of people in `poverty` remains constant.

**Poverty:** The predicted `life expectancy` increases by about 0.31
years with every 100 million person increase in the number of people in
`poverty`, when the `Human Development Index` is held constant.

### Evaluating the Model

```{r}
plot(firstmultiv, which = 1)
plot(firstmultiv, which = 2) 
```

L - Good. The residuals vs fitted plot shows us a strong linearity
between the variables since they cluster evenly about the zero line.

I - Good. The data contains entries from individual countries for the
your 2025 and do not repeat/ have duplicates. Therefore, it is
reasonable to assume that the data has no direct relationship/ influence
to each other.

N - Okay. The Q-Q plot shows the data lying closely along the line with
some dispersion at the ends of the data line. This means it is
reasonable to assume a somewhat general normal distribution among the
variables.

E - Okay. The residuals vs fitted plot show a generally good fit. There
is a bit more clustering/ tighter fit as the plot moves towards the
right. This means there is an assumption of equal variance
(Homoscedasticity), for the most part.

### Model Summary

```{r}
summary(firstmultiv)
```

#### p-value(s)

**HDI:** $p-value=2.2e-16$

Since the p-value of the `HDI` is very small (threshold: $p<=0.05$), we
can say that the slope is significantly different from 0. Therefore it
is *statistically significant*.

**Poverty:** $p-value=0.893$

Since the p-value of the number of people in `Poverty` is not considered
small (threshold: $p<=0.05$), we *cannot* say that the slope is
significantly different from 0. Therefore it is NOT *statistically
significant*.

#### $R^2$

$R^2 = 0.7499$

This means that about 74.99% of the variation in `life expectancy` is
explained by the model containing `HDI`and `Poverty`.

#### Adjusted $R^2$

$R^2_{adj}=0.7471$

The adjusted $R^2$ value accounts for additional independent variables.
Therefore, this tells us that about 74.71% of the variation in
`life expectancy` is explained by the model containing `HDI`and
`Poverty`.

#### RSE

$RSE=3.511$

This means that the "average" distance of the `life expectancy` from its
predicted value is about 3.5 years.

## Human Dev Index & GDP & Region: Multivariate Model

```{r}
secondmultiv=Full_2025|>
  lm(life_Exp~HumDevIndx+GDP+world_4region, data= _) 
secondmultiv

```

### Interpreting the Slope

*Equation:*
$\hat{y}_{LifeExp}=48.59+29.95x_{HumDevIndx}+0.00002487x_{GDP}+3.412x_{americas}+2.621x_{asia}+3.201x_{europe}$

*Slope:*

**HDI:** 29.95

**GDP:** 0.00002487

**Region_americas:** 3.412

**Region_asia:** 2.621

**Region_europe:** 3.201

**Region_africa:** This is our Categorical indicator. Note that if all
other world_4regions are 0, then `world_4regionafrica` is 1.

*Interpretation:*

**Human Dev Index:** The predicted `life expectancy` increases by about
3 years with every 0.1 unit increase in the `Human Development Index`,
assuming all other variables remain constant (GDP and region).

**GDP:** The predicted `life expectancy` increases by about 2.5 years
with every \$100,000 increase in the `GDP` per capita, assuming all
other variables remain constant (HDI and region).

**Regions:**

*americas:* The predicted `life expectancy` increases by about 3.4 years
more for countries in the `americas` region than countries in `africa`
region assuming all other variables remain constant (HDI and GDP).

*asia:* The predicted `life expectancy` increases by about 2.6 years
more for countries in the `asia` region than countries in `africa`
region assuming all other variables remain constant (HDI and GDP).

*europe:* The predicted `life expectancy` increases by about 3.2 years
more for countries in the `europe` region than countries in `africa`
region assuming all other variables remain constant (HDI and GDP).

*africa:* The predicted `life expectancy` is about 49 years if `HDI` and
`GDP` are 0 and the region is `africa`. Note, however, this is
unrealistic given the context as it does not make sense for `HDI` and
`GDP`to be 0.

### Evaluating the Model

```{r}
plot(secondmultiv, which = 1)
plot(secondmultiv, which = 2) 
```

L - Good. The residuals vs fitted plot shows us a strong linearity
between the variables since they cluster evenly about the zero line.

I - Okay. The assumption of independence is reasonably met because there
are no obvious dependencies or patterns within the data or between the
individual observations. However, we should still be aware of potential
regional similarities/ dependencies between countries that exist in the
same region.

N - Good. The Q-Q plot shows the data lying closely along the line. This
means it is reasonable to assume general normal distribution among the
variables.

E - Okay. The residuals vs fitted plot show a generally good fit. There
is a bit more clustering/ tighter fit as the plot moves towards the
right. This means there is an assumption of equal variance
(Homoscedasticity), for the most part.

### Model Summary

```{r}
summary(secondmultiv)
```

#### p-value(s)

**Human Dev Index:** $p-value=3.75e-15$

Since the p-value of the `HDI` is very small (threshold: $p<=0.05$), we
can say that the slope is significantly different from 0. Therefore it
is *statistically significant*.

**GDP:** $p-value=0.144240$

Since the p-value of the `GDP` is not considered small (threshold:
$p<=0.05$), we *cannot* say that the slope is significantly different
from 0. Therefore it is NOT *statistically significant*.

**Regions:** *NOTE:* As long as one of the categorical indicators has
slope significantly different from zero, then we include in the model.

*americas:* $p-value=0.000289$

Since the p-value of the `americas` region is very small (threshold:
$p<=0.05$), we can say that the slope for `americas` region vs `africa`
region is significantly different from zero.

*asia:* $p-value=0.001011$

Since the p-value of the `asia` region is very small (threshold:
$p<=0.05$), we can say that the slope for `asia` region vs `africa`
region is significantly different from zero.

*europe:* $p-value=0.003322$

Since the p-value of the `europe` region is very small (threshold:
$p<=0.05$), we can say that the slope for `europe` region vs `africa`
region is significantly different from zero.

#### $R^2$

$R^2=0.771$

This means that about 77.10% of the variation in `life expectancy` is
explained by the model containing `HDI`, `GDP`, and `world_4regions`.

#### Adjusted $R^2$

$R^2_{adj}=0.7645$

The adjusted $R^2$ value accounts for additional independent variables.
Therefore, this tells us that about 76.45% of the variation in
`life expectancy` is explained by the model containing `HDI`, `GDP`, and
`world_4regions`.

#### RSE

$RSE=3.388$

This means that the "average" distance of the `life expectancy` from its
predicted value is about 3.4 years.

# Comparing Models

TODO: 1. Which model is the best?

# Questions About the Data

TODO: 1. Ask a question (Should explore relationship between at least 2
variables in the dataset) 2. Answer the question 3. Create a graph/plot
that helps answer the question

*Question:*

```{r}

```

*Answer:*
