---
title: "MiniProject1"
author: "Grace Rubio, Oyinda Sanni, Mia Mason"
format: pdf
---

**Important Note** All code needs to have an answer to the "Where did u get ur code" question

# Preparing the Data
## Importing Data
TODO:
1. Import the data
```{r}

```

## Wrangling Data
TODO:
1. Combine all the data into a large dataset
```{r}

```

# Variable Report
TODO:
1. Identify each variable
2. Type of each variable
3. Information about the types of values the variable can take
4. Amount of missing data within the variable

## Cleaning Data
```{r}

```

## Summary Statistics
TODO:
1. Categorical variables: create a table of possible values and include counts or proportions.
  1a. Create an appropriate plot to show   distribution of values
2. Quantitative variables: include basic summary stats like mean, median, standard deviation, min, max, etc.
  2a. Create an appropriate plot to show distribution of values
3. Look up country involved for any extreme outliers
```{r}

```


## Variable Descriptions
TODO:
1. Write a few sentences for each variable to describe what we have learned about the variable. Incorporate information from the Summary Statistics

# Linear Models
TODO:
1. Make at least 3 linear models (1 simple, 2 multivariate)
2. Evaluate the models for LINE(Q-Q plots needed), p-values ,R2, and adjusted R2, RSE, MSE, RMSE
```{r}

```


# Comparing Models
TODO:
1. Which model is the best?

# Questions About the Data
TODO:
1. Ask a question (Should explore relationship between at least 2 variables in the dataset)
2. Answer the question
3. Create a graph/plot that helps answer the question

```{r}

```

