---
title: "MiniProject1"
author: "Grace Rubio, Oyinda Sanni, Mia Mason"
format: pdf 
editor:    
  markdown:      
    wrap: 72 
---

**Important Note** All code needs to have an answer to the "Where did u
get ur code" question

# Preparing the Data

## Load in packages

```{r}
#| label: setup 
#| include: false
#| library(tidyverse)
#| library(tidymodels)
```

## Importing Data

1.  Import the data

```{r}
# Response Variable
life_exp <-read_csv("Data/lex.csv")   
# Predictor Variables 
CO2_emissions <-read_csv("Data/co2_cons.csv")
gdp <-read_csv("Data/gdp_pcap.csv")
humanDevIndx <-read_csv("Data/hdi_human_development_index.csv")
poverty<-read_csv("Data/number_of_people_in_poverty.csv")
gapminder_2025<-read_csv("Data/gapminder_2025.csv") # QUALITATIVE (regions) 
```

## Wrangling Data

TODO: 1. Combine all the data into a large dataset

*NOTE:* We will use the last recorded year for the datasets that do not
have a 2025 entry.

```{r}
data_lst<- list(CO2_emissions, gdp, humanDevIndx, poverty, gapminder_2025)

Full_2025<-life_exp 
for (dataset in data_lst){  
  Full_2025 <- Full_2025|>
    full_join(dataset, by="geo")
}

Full_2025<-Full_2025|>
  select(geo,name.x,`2025.x`, # keep life_exp 2025 columns
         `2022.y`, # keep 2025 co2 column
         `2025.y`,  # keep 2025 gdp column
         `2023.x.x`,  # keep 2025 humanDevIdnx column
         `2025`,     # keep 2025 poverty column
         world_4region # keep 2025 regions column
         )|>
  rename(name = name.x, #rename columns for clarity
         life_Exp=`2025.x`,
         CO2=`2022.y`,
         GDP=`2025.y`,
         Poverty = `2025`,
         HumDevIndx= `2023.x.x`
         ) 
```

# Variable Report

TODO: 1. Identify each variable 2. Type of each variable 3. Information
about the types of values the variable can take 4. Amount of missing
data within the variable 5. Check if cleaning is needed and clean the
data.

## Data Cleaning

For the most part the `Full_2025` table is tidy. The main issue with the
table is with the missing values in a few of the variables. As such, we
will clean the table by removing all missing entries.

*Note:* Below, we further identify issues with each individual variable
specifically. This way, we can explicitly distinguish if/why the
cleaning performed in this section was necessary.

```{r}
# Clean the Data
Full_2025<-Full_2025|>   
  drop_na()
```

## Life Expectancy at Birth

*Type:* This variable type is quantitative.

*Info:* The Life Expectancy at Birth variable is the average number of
years a hypothetical newborn child will live if the age-specific
mortality rates observed in a country during that specific year were to
remain constant throughout their lifetime. It is an accumulation of
broader health standards, healthcare access, sanitation, nutrition, and
disease exposure in the given country.

### Summary Statistics

```{r}
#| echo: true
#| fig-align: center  

# SUMMARY
LifeExp_summary <- Full_2025|>
  summarize(avg_LifeExp = mean(life_Exp),
            sd_LifeExp = sd(life_Exp),
            min_LifeExp = min(life_Exp),
            max_LifeExp = max(life_Exp),
            mdn_LifeExp = median(life_Exp),
            iqr = IQR(life_Exp),
            sample_size = n())
LifeExp_summary

# PLOT
Full_2025|>
  ggplot(mapping = aes(x = life_Exp)) +
  geom_histogram(binwidth = 3, color = '#ACC7B4', fill = "#331B3F") +
  labs(x = "Total Life Expectancy",
       y = "Count",
       title = "Count of Country's Total Life Expectancy at Birth")
```

### Variable Descriptions

The distribution of the variable is slightly left skewed with the
majority of the countries life expectancy being concentrated between 70
and 80 years. There are no severe outliers visible in the distribution.
The minimum life expectancy is 55.48 years, and the maximum is 85.71
years. Given the results of the summary table for the variable, the
average life expectancy at birth is approximately 73.31 years, with a
standard deviation of 6.98 years. When comparing the mean to the median
of 74.12 years, this highlights the slight left skew as the countries
with shorter life expediencies pull the mean slightly below the median
value. The IQR says that the middle 50% of countries fall within a range
of about 10.29 years.

## CO2 Emissions

*Type:* This variable type is quantitative.

*Info:* This variable is the carbondioxide total (consumption based).
This variable can take numeric or NAN values. The CO2 emissions column
is measured in million tonnes of CO2. There are currently two missing
entries in the variable, so this column does require some cleaning.

### Summary Statistics (By Itself)

```{r}
#| echo: true
#| fig-align: center  

# SUMMARY
CO2_summary <- Full_2025|>
  summarize(avg_CO2 = mean(CO2),
            sd_CO2 = sd(CO2),
            min_CO2 = min(CO2),
            max_CO2 = max(CO2),
            mdn_CO2 = median(CO2),
            iqr = IQR(CO2),
            sample_size = n())
CO2_summary

# PLOT
Full_2025|>
  ggplot(mapping=aes(x = CO2)) +
  geom_histogram(binwidth = 300, color = '#ACC7B4', fill = "#331B3F") +
  labs(x = "Total CO2 Emissions (Million Tonnes)",
       y = "Count",
       title = "Count of Country's Total CO2 Emission")+

  # Add markers to identify outliers
  geom_point(data = Full_2025[Full_2025$CO2 > 2000,],
             aes(x = CO2, y = 0),
             color = "red", shape = 1, size = 12) 

```

### Summary Statistics (W/ Response Variable)

```{r}
#| echo: true
#| fig-align: center  

#SUMMARY
LifeExp_CO2_summary <- Full_2025|>
  summarize(avg_LifeExp = mean(life_Exp),
            sd_LifeExp = sd(life_Exp),
            min_LifeExp = min(life_Exp),
            max_LifeExp = max(life_Exp),
            mdn_LifeExp = median(life_Exp),
            iqr_LifeExp = IQR(life_Exp),
            
            avg_CO2 = mean(CO2),
            sd_CO2 = sd(CO2),
            min_CO2 = min(CO2),
            max_CO2 = max(CO2),
            mdn_CO2 = median(CO2),
            iqr_CO2 = IQR(CO2),
            
            sample_size = n())
LifeExp_CO2_summary

#PLOT
Full_2025|>
  ggplot(aes(x = CO2, y = life_Exp)) +
  geom_point() +
  labs(x = "CO2 emissions (Million Tonnes)",
       y = "Life Expectancy at Birth",
       title = "Life expectancy at Birth vs. CO2 Emissions")
```

### Variable Descriptions

We can see that the distribution of total CO2 emissions across the
listed countries is right-skewed with the majority of countries falling
before the 1 billion tonnes mark. There are two main outliers of this
variable (India and the US) that likely have an impact on our summary
results, so that should be taken into account with later analysis. Given
the results of our summary for this variable, we can see that the
average CO2 emission per country is about 137.15 million tonnes. By
comparing the mean to our median (17.7 million tonnes), the
right-skewness of the data becomes more apparent as the difference
between the average CO2 emission and the middle value further suggests
the outliers are pulling the mean up. The IQR is helpful for this skewed
distribution, as it tells us that the middle 50% of countries have CO₂
emissions within a range of about 70 million tonnes.

## GDP per Capita

*Type:* This variable type is quantitative.

*Info:* GDP per capita is the gross domestic product which is the total
moetary value of all goods and services produced in a country in a given
year. The values in this dataset are stored as float values. GDP per
capita is recorded in US dollars, specifically PPP\$ or purchasing power
parity, and values are adjusted for inflation.

### Summary Statistics (By Itself)

```{r}
#| echo: true
#| fig-align: center

# SUMMARY
GDP_summary <- Full_2025|>
  summarize(avg_GDP = mean(GDP),
            sd_GDP = sd(GDP),
            min_GDP = min(GDP),
            max_GDP = max(GDP),
            mdn_GDP = median(GDP),
            iqr = IQR(GDP),
            sample_size = n())
GDP_summary

# PLOT
Full_2025|>
  ggplot(mapping=aes(x = GDP)) +
  geom_histogram(binwidth = 3000, color = '#ACC7B4', fill = "#331B3F") +
  labs(x = "Total GDP per Capita (Dollars)",
       y = "Count",
       title = "Count of Country's Total GDP per Capita")+
# Add markers to identify outliers
  geom_point(data = Full_2025[Full_2025$GDP > 90000,],
             aes(x = GDP, y = 0),
             color = "red", shape = 1, size = 12)
```

### Summary Statistics (W/ Response Variable)

```{r}
#| echo: true
#| fig-align: center  

#SUMMARY
LifeExp_GDP_summary <- Full_2025|>
  summarize(avg_LifeExp = mean(life_Exp),
            sd_LifeExp = sd(life_Exp),
            min_LifeExp = min(life_Exp),
            max_LifeExp = max(life_Exp),
            mdn_LifeExp = median(life_Exp),
            iqr_LifeExp = IQR(life_Exp),
            
            avg_GDP = mean(GDP),
            sd_GDP = sd(GDP),
            min_GDP = min(GDP),
            max_GDP = max(GDP),
            mdn_GDP = median(GDP),
            iqr_GDP = IQR(GDP),
            
            sample_size = n())
LifeExp_GDP_summary

#PLOT
Full_2025|>
  ggplot(aes(x = GDP, y = life_Exp)) +
  geom_point() +
  labs(x = "GDP per Capita (US$)",
       y = "Life Expectancy at Birth",
       title = "Life expectancy at Birth vs. GDP per Capita")
```

### Variable Descriptions

The distribution of GDP per capita across the listed countries in the
Full_2025 dataset is concentrated on the right side of the graph, with a
majority of it being below \$25,000. There are four outliers which reach
up to around \$140,000, causing the mean to be higher in the summary.
Given the summary, the average GDP per capita is \~\$24,859, which
comparing to the median of \$15,343 highlights the right-sided
concentration. The IQR of \$32,353 provides a better sense of spread for
this distribution, indicating that the middle 50% of countries fall
within a range of roughly \$32,353.

## Human Development Index

*Type:* This variable type is quantitative.

*Info:* The HumDevIndx variable is the measure of overall human
well-being and development of a country. It combines three key
dimensions of human development like health or longevity, education, and
standard of living. These area measured by life expectancy at birth,
mean years of schooling for adults and expected years of schooling for
children, and the gross national income per capita respectively. The
variable ranges from 0 to 1, where 1 indicates higher levels of human
development, and is a float type.

### Summary Statistics (By Itself)

```{r}
#| echo: true
#| fig-align: center

# SUMMARY
HumDevIndx_summary <- Full_2025|>
  summarize(avg_HDI = mean(HumDevIndx),
            sd_HDI = sd(HumDevIndx),
            min_HDI = min(HumDevIndx),
            max_HDI = max(HumDevIndx),
            mdn_HDI = median(HumDevIndx),
            iqr = IQR(HumDevIndx),
            sample_size = n())
HumDevIndx_summary

# PLOT
Full_2025|>
  ggplot(mapping=aes(x = HumDevIndx)) +
  geom_histogram(binwidth = 0.02, color = '#ACC7B4', fill = "#331B3F") +
  labs(x = "Total Human Development Index",
       y = "Count",
       title = "Count of Country's Total Human Development Index")
```

### Summary Statistics (W/ Response Variable)

```{r}
#| echo: true
#| fig-align: center  

#SUMMARY
LifeExp_HDI_summary <- Full_2025|>
  summarize(avg_LifeExp = mean(life_Exp),
            sd_LifeExp = sd(life_Exp),
            min_LifeExp = min(life_Exp),
            max_LifeExp = max(life_Exp),
            mdn_LifeExp = median(life_Exp),
            iqr_LifeExp = IQR(life_Exp),
            
            avg_HDI = mean(HumDevIndx),
            sd_HDI = sd(HumDevIndx),
            min_HDI = min(HumDevIndx),
            max_HDI = max(HumDevIndx),
            mdn_HDI = median(HumDevIndx),
            iqr_HDI = IQR(HumDevIndx),
            
            sample_size = n())
LifeExp_HDI_summary

#PLOT
Full_2025|>
  ggplot(aes(x = HumDevIndx, y = life_Exp)) +
  geom_point() +
  labs(x = "Human Development Index",
       y = "Life Expectancy at Birth",
       title = "Life expectancy at Birth vs. CO2 Emissions")
```

### Variable Descriptions

The distribution of the HDI scores across the countries is concentrated
to the left, with the majority of them having an HDI score above 0.7.
There are no extreme outliers in the distribution due to the variable
being bound to a scale with a minimum of 0.388 and a maximum of 0.972.
Given the results of the summary for the variable, the average HDI score
across countries is approximately 0.734. Comparing the mean to the
median of 0.75, the slight left skew becomes apparent, as the middle
value sits slightly higher than the average due to the tail of lower HDI
scores bringing the mean down. The IQR shows that the middle 50% of
countries fall within a narrow range of about 0.236.

## People in Poverty

*Type:* This variable type is quantitative.

*Info:* The poverty variable records the estimated population in a
country whose daily income or consumption falls before the international
extreme poverty threshold. The unit of measure is number of people.

### Summary Statistics (By Itself)

```{r}
#| echo: true
#| fig-align: center

# SUMMARY
Poverty_summary <- Full_2025|>
  summarize(avg_Poverty = mean(Poverty),
            sd_Poverty = sd(Poverty),
            min_Poverty = min(Poverty),
            max_Poverty = max(Poverty),
            mdn_Poverty = median(Poverty),
            iqr = IQR(Poverty),
            sample_size = n()) 
Poverty_summary

# PLOT
Full_2025|>   
  ggplot(mapping=aes(x = Poverty)) +
  geom_histogram(binwidth = 3000000, color = '#ACC7B4', fill = "#331B3F") +
  labs(x = "Total Poverty Population (Millions of People)",
       y = "Count",
       title = "Count of Country's Total Poverty Population")+

# Add markers to identify outliers
  geom_point(data = Full_2025[Full_2025$Poverty > 40000000,],
             aes(x = Poverty, y = 0),
             color = "red", shape = 1, size = 12)
```

### Summary Statistics (W/ Response Variable)

```{r}
#| echo: true
#| fig-align: center  

#SUMMARY
LifeExp_Poverty_summary <- Full_2025|>
  summarize(avg_LifeExp = mean(life_Exp),
            sd_LifeExp = sd(life_Exp),
            min_LifeExp = min(life_Exp),
            max_LifeExp = max(life_Exp),
            mdn_LifeExp = median(life_Exp),
            iqr_LifeExp = IQR(life_Exp),
            
            avg_Poverty = mean(Poverty),
            sd_Poverty = sd(Poverty),
            min_Poverty = min(Poverty),
            max_Poverty = max(Poverty),
            mdn_Poverty = median(Poverty),
            iqr_Poverty = IQR(Poverty),
            
            sample_size = n())
LifeExp_Poverty_summary

#PLOT
Full_2025|>
  ggplot(aes(x = Poverty, y = life_Exp)) +
  geom_point() +
  labs(x = "Ppulation of Poverty",
       y = "Life Expectancy at Birth",
       title = "Life expectancy at Birth vs. CO2 Emissions")
```

### Variable Descriptions

The distribution of the total poverty population across the countries is
very right-skewed, with the majority of them being clustered below 10
million. There are four distinct outliers to the right with the maximum
reaching a population as high as 95,605,313 people. This maximum value
strongly influences the results of the summary table and should be taken
into account for the analysis. From the summary results, the average
poverty population per country is 4,615,385 people, and a very high
standard deviation of 12,475,121. Comparing the mean to the median of
182,163 people, the extreme right skew is very evident and shows how the
few high-poverty countries inflate the mean. The IQR shoes that the
middle 50% of countries have a poverty population spanning a range of
about 2.42 million people.

## Region

*Type:* This variable type is categorical.

*Info:* The region variable stores information about the continent in
which the country is located in. The variable can be one of the four
following values: americas (includes north and south), africa, asia, and
europe.

### Summary Statistics (By Itself)

```{r}
#| echo: true
#| fig-align: center

# SUMMARY
region_summary <- Full_2025|>
  group_by(world_4region) |>
  summarize(count = n())
region_summary  

# PLOT 
Full_2025|>
  ggplot(mapping=aes(x = world_4region)) +
  geom_bar(color = '#ACC7B4', fill = "#331B3F") +
  labs(x = "World Regions",
       y = "Count", 
       title = "Distribution of Countries Across World Regions")
```

### Summary Statistics (W/ Response Variable)

```{r}
#| echo: true
#| fig-align: center  

#SUMMARY
LifeExp_region_summary <- Full_2025|>
  group_by(world_4region) |>
  summarize(avg_LifeExp = mean(life_Exp),
          sd_LifeExp = sd(life_Exp),
          min_LifeExp = min(life_Exp),
          max_LifeExp = max(life_Exp),
          mdn_LifeExp = median(life_Exp),
          iqr_LifeExp = IQR(life_Exp),
          count = n())
LifeExp_region_summary

#PLOT
Full_2025|>
  ggplot(aes(x = world_4region, y = life_Exp)) +
  geom_boxplot() +
  labs(x = "World Regions",
       y = "Life Expectancy at Birth",
       title = "Life expectancy at Birth vs. World Regions")
```

### Variable Descriptions

Africa and Asia are the most represented regions out of the countries in
the dataset with their counts being 53 and 52 respectively. The Americas
are the least represented world region being the minimum at 34 total
countries.

# Linear Models

TODO: 1. Make at least 3 linear models (1 simple, 2 multivariate) 2.
Evaluate the models for LINE(Q-Q plots needed), p-values ,R2, and
adjusted R2, RSE, MSE, RMSE

## GDP per Capita: Simple Linear Model

```{r}
lifegdp=Full_2025|>
  lm(life_Exp~GDP, data= _) 
lifegdp
```

### Interpreting the Slope

*Equation:* $\hat{y}_{LifeExp}=68.51+0.0001935x_{GDP}$

*Slope:* 0.0001935

*Interpretation:* The slope of this model tells us that For every \$1
increase in gdp per capita there will be a 0.0001935 year increase in
predicted life expectancy. In other words, the predicted
`life expectancy` of a country increases by 1.9 years for every \$10,000
increase in `GDP`

### Evaluating the Model

```{r}
plot(lifegdp, which = 1)
plot(lifegdp, which = 2) 
```

L - Bad. The residuals vs fitted plot shows us a poor linearity between
the variables. We can see this because the points do not cluster/ evenly
distribute around zero. We can aslo see the line curve downwards as the
fitted values increase.

I - Not good. There are patterns in the geographic location/region of
the data. As such, we cannot assume independence.

N - Okay. The Q-Q plot shows the data lying closely along the line for
(the most part). While there is some minor divergence (especially
towards the end), we can still say that it is reasonable to assume
general normal distribution among the variables.

E - Bad. The variance declines as fitted value increases (i.e., the
variance is NOT constant).

### Model Summary

```{r}
summary(lifegdp)
```

#### p-value

$p-value=2.2e-16$

Since the p-value is very small (threshold: $p<=0.05$), we can say that
the slope is significantly different from 0. Therefore it is
*statistically significant*.

#### $R^2$

$R^2 = 0.5041$

50.41% of the variation of life expectancy is explained by the linear
model including gdp per capita.

#### RSE

$RSE = 4.93$

The "average" distance of the life expectancy from its predicted value
is about 5 years.

## Poverty & Human Development Index: Multivariate Model

```{r}
firstmultiv=Full_2025|>
  lm(life_Exp~HumDevIndx+Poverty, data= _) 
firstmultiv
```

### Interpreting the Slope

*Equation:*
$\hat{y}_{LifeExp}=44.07+39.85x_{HumDevIndx}+0.000000003099x_{Poverty}$

*Slope:*

**Human Dev Index:** 39.85

**Poverty:** 0.000000003099

*Interpretation:*

**Human Dev Index:** The predicted `life expectancy` increases by about
4 years with every 0.1 unit increase in the `Human Development Index`,
when the number of people in `poverty` remains constant.

**Poverty:** The predicted `life expectancy` increases by about 0.31
years with every 100 million person increase in the number of people in
`poverty`, when the `Human Development Index` is held constant.

### Evaluating the Model

```{r}
plot(firstmultiv, which = 1)
plot(firstmultiv, which = 2) 
```

L - Good. The residuals vs fitted plot shows us a strong linearity
between the variables since they cluster evenly about the zero line.

I - Good. The data contains entries from individual countries for the
your 2025 and do not repeat/ have duplicates. Therefore, it is
reasonable to assume that the data has no direct relationship/ influence
to each other.

N - Okay. The Q-Q plot shows the data lying closely along the line with
some dispersion at the ends of the data line. This means it is
reasonable to assume a somewhat general normal distribution among the
variables.

E - Okay. The residuals vs fitted plot show a generally good fit. There
is a bit more clustering/ tighter fit as the plot moves towards the
right. This means there is an assumption of equal variance
(Homoscedasticity), for the most part.

### Model Summary

```{r}
summary(firstmultiv)
```

#### p-value(s)

**HDI:** $p-value=2.2e-16$

Since the p-value of the `HDI` is very small (threshold: $p<=0.05$), we
can say that the slope is significantly different from 0. Therefore it
is *statistically significant*.

**Poverty:** $p-value=0.893$

Since the p-value of the number of people in `Poverty` is not considered
small (threshold: $p<=0.05$), we *cannot* say that the slope is
significantly different from 0. Therefore it is NOT *statistically
significant*.

#### $R^2$

$R^2 = 0.7499$

This means that about 74.99% of the variation in `life expectancy` is
explained by the model containing `HDI`and `Poverty`.

#### Adjusted $R^2$

$R^2_{adj}=0.7471$

The adjusted $R^2$ value accounts for additional independent variables.
Therefore, this tells us that about 74.71% of the variation in
`life expectancy` is explained by the model containing `HDI`and
`Poverty`.

#### RSE

$RSE=3.511$

This means that the "average" distance of the `life expectancy` from its
predicted value is about 3.5 years.

## Human Dev Index & GDP & Region: Multivariate Model

```{r}
secondmultiv=Full_2025|>
  lm(life_Exp~HumDevIndx+GDP+world_4region, data= _) 
secondmultiv
```

### Interpreting the Slope

*Equation:*
$\hat{y}_{LifeExp}=48.59+29.95x_{HumDevIndx}+0.00002487x_{GDP}+3.412x_{americas}+2.621x_{asia}+3.201x_{europe}$

*Slope:*

**HDI:** 29.95

**GDP:** 0.00002487

**Region_americas:** 3.412

**Region_asia:** 2.621

**Region_europe:** 3.201

**Region_africa:** This is our Categorical indicator. Note that if all
other world_4regions are 0, then `world_4regionafrica` is 1.

*Interpretation:*

**Human Dev Index:** The predicted `life expectancy` increases by about
3 years with every 0.1 unit increase in the `Human Development Index`,
assuming all other variables remain constant (GDP and region).

**GDP:** The predicted `life expectancy` increases by about 2.5 years
with every \$100,000 increase in the `GDP` per capita, assuming all
other variables remain constant (HDI and region).

**Regions:**

*americas:* The predicted `life expectancy` increases by about 3.4 years
more for countries in the `americas` region than countries in `africa`
region assuming all other variables remain constant (HDI and GDP).

*asia:* The predicted `life expectancy` increases by about 2.6 years
more for countries in the `asia` region than countries in `africa`
region assuming all other variables remain constant (HDI and GDP).

*europe:* The predicted `life expectancy` increases by about 3.2 years
more for countries in the `europe` region than countries in `africa`
region assuming all other variables remain constant (HDI and GDP).

*africa:* The predicted `life expectancy` is about 49 years if `HDI` and
`GDP` are 0 and the region is `africa`. Note, however, this is
unrealistic given the context as it does not make sense for `HDI` and
`GDP`to be 0.

### Evaluating the Model

```{r}
plot(secondmultiv, which = 1)
plot(secondmultiv, which = 2) 
```

L - Good. The residuals vs fitted plot shows us a strong linearity
between the variables since they cluster evenly about the zero line.

I - Okay. The assumption of independence is reasonably met because there
are no obvious dependencies or patterns within the data or between the
individual observations. However, we should still be aware of potential
regional similarities/ dependencies between countries that exist in the
same region.

N - Good. The Q-Q plot shows the data lying closely along the line. This
means it is reasonable to assume general normal distribution among the
variables.

E - Okay. The residuals vs fitted plot show a generally good fit. There
is a bit more clustering/ tighter fit as the plot moves towards the
right. This means there is an assumption of equal variance
(Homoscedasticity), for the most part.

### Model Summary

```{r}
summary(secondmultiv)
```

#### p-value(s)

**Human Dev Index:** $p-value=3.75e-15$

Since the p-value of the `HDI` is very small (threshold: $p<=0.05$), we
can say that the slope is significantly different from 0. Therefore it
is *statistically significant*.

**GDP:** $p-value=0.144240$

Since the p-value of the `GDP` is not considered small (threshold:
$p<=0.05$), we *cannot* say that the slope is significantly different
from 0. Therefore it is NOT *statistically significant*.

**Regions:** *NOTE:* As long as one of the categorical indicators has
slope significantly different from zero, then we include in the model.

*americas:* $p-value=0.000289$

Since the p-value of the `americas` region is very small (threshold:
$p<=0.05$), we can say that the slope for `americas` region vs `africa`
region is significantly different from zero.

*asia:* $p-value=0.001011$

Since the p-value of the `asia` region is very small (threshold:
$p<=0.05$), we can say that the slope for `asia` region vs `africa`
region is significantly different from zero.

*europe:* $p-value=0.003322$

Since the p-value of the `europe` region is very small (threshold:
$p<=0.05$), we can say that the slope for `europe` region vs `africa`
region is significantly different from zero.

#### $R^2$

$R^2=0.771$

This means that about 77.10% of the variation in `life expectancy` is
explained by the model containing `HDI`, `GDP`, and `world_4regions`.

#### Adjusted $R^2$

$R^2_{adj}=0.7645$

The adjusted $R^2$ value accounts for additional independent variables.
Therefore, this tells us that about 76.45% of the variation in
`life expectancy` is explained by the model containing `HDI`, `GDP`, and
`world_4regions`.

#### RSE

$RSE=3.388$

This means that the "average" distance of the `life expectancy` from its
predicted value is about 3.4 years.

# Comparing Models

TODO: 1. Which model is the best?

The Human Development Index, GDP, and Region model performed te best. Its adjusted R^2 is 0.7645, this means there is a 0.7645 variation in the life expectancy explained in the variables after we accounted for the predictors. It also has the lowest RSE of 3.888 so the prediction are fairly close to the actual life expectancy.
The simple GDP model performed the worst because it only explains about 0.5041 of teh variation in life expectancy and has the highest RSE of 4.93 years. The HDI and poverty model did better, but it has a lower adjusted R^2 of 0.7471 and it RSE is slighly higher than in the The Human Development Index, GDP, and Region model of 3.511.
Therefore, based on adjusted R², RSE, and the LINE assumptions, the Human Development Index, GDP, and Region model provides the best fit.

# Questions About the Data

TODO: 1. Ask a question (Should explore relationship between at least 2
variables in the dataset) 2. Answer the question 3. Create a graph/plot
that helps answer the question

*Does the relationship between Human Development Index and life expectancy differ across world regions?*

```{r}
Full_2025 |>
  ggplot(aes(x = HumDevIndx, y = life_Exp, color = world_4region)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Human Development Index",
    y = "Life Expectancy at Birth",
    color = "World Region",
    title = "Life Expectancy vs. Human Development Index by Region"
  )
*There is a positive relationship between Human Development Index and life expectancy.Countries with higher HDI values tend to have higher life expectancy. The graph also shows differences between regions. Countries within the same HDI range can have different life expectancies depending on their region. This supports the results from our multivariate model, where HDI was more and region also contributed to explaining differences in life expectancy.*

```
